Computing in the Field of Rationals

### 2012-06-26 rational_ext.tex
start of new intro to section 2

@@

@@

\newcommand{\arrow}[1]{\mathtt{#1}}

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{\asterix}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! [^ = \lceil
%! ^] = \rceil
%! {| = \{
%! |} = \}
%! dots = \ldots
%! ^^^ = ^{\dagger}
%! elem = \in
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! Pi = \Pi
%! Pi0 = \Pi^{o}
%! PiEE* = \Pi^{\eta\epsilon}_{*}
%! PiEE+ = \Pi^{\eta\epsilon}_{+}
%! PiEE = \Pi^{\eta\epsilon}
%! CatSet = \textbf{Set}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

\title{Negative and Fractional Types}

\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Zachary Sparks} 
           {Indiana University}
           {zasparks@indiana.edu}
\authorinfo{Jacques Carette} 
           {McMaster University}
           {carette@mcmaster.ca}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}

@@

  %% Every functional programmer knows about sum and product types, {{a+b}} and
  %% {{a*b}} respectively. Negative and fractional types, {{a-b}} and {{a/b}}
  %% respectively, are much less known and their computational interpretation is
  %% unfamiliar and often complicated. We show that in a programming model in
  %% which information is preserved (such as the model introduced in our recent
  %% paper on \emph{Information Effects}), these types have particularly natural
  %% computational interpretations. Intuitively, values of negative types are
  %% values that flow ``backwards'' to satisfy demands and values of fractional
  %% types are values that impose constraints on their context.  The combination
  %% of these negative and fractional types enables greater flexibility in
  %% programming by breaking global invariants into local ones that can be
  %% autonomously satisfied by a subcomputation. Theoretically, these types give
  %% rise to \emph{two} function spaces and to \emph{two} notions of
  %% continuations, suggesting that the previously observed duality of
  %% computation conflated two orthogonal notions: an additive duality that
  %% corresponds to backtracking and a multiplicative duality that corresponds
  %% to constraint propagation.

\end{abstract}

@@

\begin{verbatim}
Restructure this section to emphasize the following

In plain Pi: 
------------

types b ::= 0 | 1 | b+b | b*b
values v ::= () | L v | R v | (v,v)
typing v : b

Semantics of values: 
 - a type b is interpreted as a set
 - a value v : b is an element of the set denoted by b

types b <-> b
combinators c ::= primitiveIso
               | c + c | c * c | c ; c 
typing c : b1 <-> b2

Semantics of combinators:
 - types b1 <-> b2 are interpreted as bijections 
   between the sets denoted by b1 and b2
 - a combinator c : b1 <-> b2 is interpeted as a 
   bijection between the appropriate sets

Evaluation is function application, i.e., 
  eval c v applies the function denoted by c to 
  the value denoted by v.

[If we want to use categorical language, 
we can express the above semantics using the category 
of finite sets and bijections.]

\end{verbatim}

We review our reversible language {{Pi}}: the presentation in this
section differs from the one in our previous paper~\cite{infeffects} in two
aspects. First, we add the empty type {{0}} which is necessary to express the
additive duality. Second, instead of explaining evaluation using a natural
semantics, we give a small-step operational semantics that is more
appropriate for the connections with continuations explored in this paper.

@@

Unlike the traditional situation with the $\lambda$-calculus, in {{Pi}},
there is a sharp distinction between data and programs.

\paragraph*{Data.} The sets of values and their types include:
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b + b | b * b 
% values, v ::= () | left v | right v | (v, v)
Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
and products types {{b1*b2}}.  Values includes {{()}} which is the only value
of type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. There are no values of
type {{0}}.

\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
{{v \rightarrow v}}. Specifically, they are witnesses to the following type
isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%zeroe :&  0 + b <-> b &: zeroi
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist0 :& 0 * b <-> 0 &: factor0
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces one or two combinators that witness
the isomorphism in the middle. Collectively the isomorphisms state that the
structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
multiplication distributes over addition. The isomorphisms are extended to
form a congruence relation by adding the following constructors that witness
equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

@@

%%%%%%%%%%%%
\subsection{Graphical Language}

@@

%%%%%%%%%%%%
\subsection{Semantics}

@@

%%%%%%%%%%%%%%%
\subsection{Constructions}

@@

%%
\paragraph*{Cloning.}
Although cloning is generally not allowed in reversible languages, it is
possible at the cost of having additional constant inputs. For example,
consider the gate {{else_{not} }}. Generally, the gate maps
{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
on the cases in which the second input is {{true}}, we get that the gate maps
{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
i.e., the gate clones the first input. A circuit of {{n}} parallel
{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
{{true}} inputs in the process.  Let us call this construction
{{clone^n_{bool} }}.

%%%%%%%%%%%%%%%%
\subsection{Categorical Structure}

@@

%%%%%%%%%%%%
\subsection{Small Step Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types : {{PiEE*}} }
\label{sec:int}

In arithmetic the main identity satisfied by fractional numbers is that for
non-zero {{b}}, we have {{b * 1/b = 1}}. Translating this identity to our
world requires us to find an isomorphism between the types {{1}} and 
{{b * 1/b}} (where {{b}} is not equivalent to {{0}}). Formally, we move 
from the setting of symmetric monoidal categories to that of compact
closed categories.

%%%%%%%%%%%%%%%%%
\subsection{The Powerset Monad}

In order to interpret fractional types, we first define a monad {{S}} over
the {{Pi}}-category of finite sets and bijections.

\paragraph*{Data.} The syntax of {{Pi}} is extended with the type {{S b}}
which denotes sets {{ {| v1, v2, dots |} }} of values of type {{b}}:
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b + b | b * b | S b
% values, v ::= () | left v | right v | (v, v) | {| v1, v2, dots |}

Types include the empty type {{0}}, the unit type {{1}}, sum types {{b1+b2}},
and products types {{b1*b2}}.  Values includes {{()}} which is the only value
of type {{1}}, {{left v}} and {{right v}} which inject {{v}} into a sum type,
and {{(v1,v2)}} which builds a value of product type. There are no values of
type {{0}}.

\paragraph*{Programs.} All the programs of {{Pi}} are special maps of type
{{v -> v}}. Specifically, they are witnesses to the following type
isomorphisms:
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%zeroe :&  0 + b <-> b &: zeroi
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist0 :& 0 * b <-> 0 &: factor0
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces one or two combinators that witness
the isomorphism in the middle. Collectively the isomorphisms state that the
structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
multiplication distributes over addition. The isomorphisms are extended to
form a congruence relation by adding the following constructors that witness
equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

%subcode{opsem} include main
% eta* &: 1 <-> (1/b) * b :& eps*

IntuitivelyThe main challenge Adding fractional types to
{{Pi}} takes us from a symmetric monoidal category to a compact closed
category.

\paragraph*{Traces and the Int-Construction.}
For a traced symmetric monoidal category, the Int-construction of
Joyal, Street and Verity \cite{joyal1996traced} gives us a way of
constructing a compact closed category. This is the obvious first
thing to try for a duality in this setting.

@@

@@

%%%%%%%%%%%%
\subsection{Semantics}

@@

%%%%%%%%%%%
\subsection{Categorical Structure}

@@

%%%%%%%%%
\subsection{Categorical Constructions}

@@

%%%%%%%
\subsection{Construction of the Solver}

@@

%%%%%%%%
\subsection{Final Details}
\label{sub:f}

@@

%%%%%%%%%
\subsection{Semantics}

@@

%%%%%%%%%%%%
\subsection{Constructions with Negatives and Fractionals}
\label{sec:specific-constructions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Two Dualities of Computation}
\label{sub:hof}

@@

%%%%%%%%%%%%%
\subsection{Function Spaces}

>>>>

%% I am getting an error about too many math packages used!!!
%% I commented the ones we don't seem to be using.

@@

%% \newtheorem{theorem}{Theorem}[section]
%% \newtheorem{lemma}[theorem]{Lemma}
%% \newtheorem{proposition}[theorem]{Proposition}
%% \newtheorem{corollary}[theorem]{Corollary}

@@

\newcommand{\arrow}[1]{\mathtt{#1}}
\newcommand{\PiL}{\ensuremath{\Pi}}

@@

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{\asterix}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <><> = \approx
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o+ = \multimap^{+}
%! -o* = \multimap^{\times}
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! [^ = \lceil
%! ^] = \rceil
%! {| = \{
%! |} = \}
%! ^^^ = ^{\dagger}
%! elem = \in
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! Pi = \Pi
%! Pi0 = \Pi^{o}
%! PiEE* = \Pi^{\eta\epsilon}_{*}
%! PiEE+ = \Pi^{\eta\epsilon}_{+}
%! PiEE = \Pi^{\eta\epsilon}
%! CatSet = \textbf{Set}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

\title{Computing in the Field of Rationals}

\authorinfo{Nicolas Bourbaki}

%% \authorinfo{Roshan P. James}
%%            {Indiana University}
%%            {rpjames@indiana.edu}
%% \authorinfo{Amr Sabry}
%%            {Indiana University}
%%            {sabry@indiana.edu}

@@

  %% Every functional programmer knows about sum and product types, {{a+b}} and
  %% {{a*b}} respectively. Negative and fractional types, {{a-b}} and {{a/b}}
  %% respectively, are much less known and their computational interpretation is
  %% unfamiliar and often complicated. We show that in a programming model in
  %% which information is preserved (such as the model introduced in our recent
  %% paper on \emph{Information Effects}), these types have particularly natural
  %% computational interpretations. Intuitively, values of negative types are
  %% values that flow ``backwards'' to satisfy demands and values of fractional
  %% types are values that impose constraints on their context.  The combination
  %% of these negative and fractional types enables greater flexibility in
  %% programming by breaking global invariants into local ones that can be
  %% autonomously satisfied by a subcomputation. Theoretically, these types give
  %% rise to \emph{two} function spaces and to \emph{two} notions of
  %% continuations, suggesting that the previously observed duality of
  %% computation conflated two orthogonal notions: an additive duality that
  %% corresponds to backtracking and a multiplicative duality that corresponds
  %% to constraint propagation.
\end{abstract}

@@

We review our reversible language {{Pi}}: the presentation in this
section differs from the one in~\cite{infeffects} in two technical
aspects. First, we add the empty type {{0}} which is necessary to express the
additive duality. Second, instead of explaining evaluation using a natural
semantics, we give a small-step operational semantics that is more
appropriate for later parts of this paper.

There is also a notable shift in emphasis.  The terms of {{Pi}} are not
classical values and functions; rather, the terms are isomorphism 
witnesses.  In other words, the terms of {{Pi}} are proofs that certain
``shapes of values'' are isomorphic.  And, in classical Curry-Howard
fashion, our operational semantics shows how these proofs can be
directly interpreted as actions on ordinary values which effect this
shape transformation.

Of course, ``shapes of values'' are very familiar already: they are usually
called \emph{types}.  But usually one designs a type system as a method
of classifying terms, with the eventual purpose to show that certain
properties of well-typed terms hold, such as safety.  Our eventual goal
is different: we start from a type system, and are striving to discover
a term language which naturally inhabits these types, along with an
appropriate operational semantics.

Generally, such an approach would be rather vacuous.  What makes this
pursuit different is that ... (will continue from here -- Jacques)

@@

The set of types includes the empty type {{0}}, the unit type {{1}}, sum
types {{b1+b2}}, and products types {{b1*b2}}. The set of values {{v}}
includes {{()}} which is the only value of type {{1}}, {{left v}} and {{right v}} 
which inject {{v}} into a sum type, and {{(v1,v2)}} which builds a
value of product type. There are no values of type {{0}}:
%subcode{bnf} include main
% value types, b ::= 0 | 1 | b + b | b * b 
% values, v ::= () | left v | right v | (v, v)

The combinators of {{Pi}} are witnesses to the following type
isomorphisms: 
%subcode{bnf} include main
%! columnStyle = r@{\hspace{-0.5pt}}c@{\hspace{-0.5pt}}l
%zeroe :&  0 + b <-> b &: zeroi
%swap+ :&  b1 + b2 <-> b2 + b1 &: swap+
%assocl+ :&  b1 + (b2 + b3) <-> (b1 + b2) + b3 &: assocr+
%identl* :&  1 * b <-> b &: identr*
%swap* :&  b1 * b2 <-> b2 * b1 &: swap*
%assocl* :&  b1 * (b2 * b3) <-> (b1 * b2) * b3 &: assocr*
%dist0 :& 0 * b <-> 0 &: factor0
%dist :&~ (b1 + b2) * b3 <-> (b1 * b3) + (b2 * b3)~ &: factor
Each line of the above table introduces one or two combinators that witness
the isomorphism in the middle. Collectively the isomorphisms state that the
structure {{(b,+,0,*,1)}} is a \emph{commutative semiring}, i.e., that each
of {{(b,+,0)}} and {{(b,*,1)}} is a commutative monoid and that
multiplication distributes over addition. The isomorphisms are extended to
form a congruence relation by adding the following constructors that witness
equivalence and compatible closure:
%subcode{proof} include main
%@  ~
%@@ id : b <-> b 
%
%@ c : b1 <-> b2
%@@ sym c : b2 <-> b1
%
%@ c1 : b1 <-> b2
%@ c2 : b2 <-> b3
%@@ c1(;)c2 : b1 <-> b3
%---
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (+) c2 : b1 + b2 <-> b3 + b4
%
%@ c1 : b1 <-> b3
%@ c2 : b2 <-> b4
%@@ c1 (*) c2 : b1 * b2 <-> b3 * b4

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Graphical Language}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions}

@@

\paragraph*{Cloning.}
Although cloning is generally not allowed in reversible languages, it is
possible at the cost of having additional constant inputs. For example,
consider the gate {{else_{not} }}. Generally, the gate maps
{{(false,a)}} to {{(false,not a)}} and {{(true,a)}} to {{(true,a)}}. Focusing
on the cases in which the second input is {{true}}, we get that the gate maps
{{(false,true)}} to {{(false,false)}} and {{(true,true)}} to {{(true,true)}},
i.e., the gate clones the first input. A circuit of {{n}} parallel
{{else_{not} }} gates can hence clone {{n}} bits.  They also consume {{n}}
{{true}} inputs in the process.  Let us call this construction
{{clone^n_{bool} }}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Categorical Structure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Small Step Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{First class functions from the permutation space.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Traces and the Int-Construction.}
\label{sec:int}

For a traced symmetric monoidal category, the Int-construction of
Joyal, Street and Verity \cite{joyal1996traced} gives us a way of
constructing a compact closed category. This is the obvious first
thing to try for a duality in this setting.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types : {{PiEE*}} }

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Categorical Structure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Categorical Constructions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Construction of the Solver}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Final Details}
\label{sub:f}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Constructions with Negatives and Fractionals}
\label{sec:specific-constructions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Two Dualities of Computation}
\label{sub:hof}

@@

\subsection{Function Spaces}
